\documentclass[12pt,fleqn]{extarticle}


% Enable a table of contents
\setcounter{tocdepth}{2}   % sections and subsections
\setcounter{secnumdepth}{2}



% --- Page layout ---

\usepackage{setspace}        % Optional: adjust line spacing
\setstretch{1.08}
\usepackage[margin=0.75in]{geometry}
% --- Fonts ---
%\usepackage{tgtermes}        % Modern serif text font (TeX Gyre Termes)
\usepackage{newtxtext}
\usepackage{newtxmath}       % Matching math font with good coverage


\usepackage{enumitem}
\setlist[itemize]{itemsep=2pt, topsep=2pt, parsep=0pt, partopsep=0pt}



% --- General utilities ---
%\usepackage{amsmath}
\usepackage{mathtools}

\usepackage{graphicx}
\usepackage{xcolor}
\usepackage{upgreek}

%\newcommand{\ttq}[1]{\texttt{`#1'}}
\newcommand{\ttq}[1]{\texttt{\textquoteleft#1\textquoteright}}
\usepackage{csquotes}


%\DeclareRobustCommand{\funconvert}{\texttt{function\kern0.21em convert}\hspace{0.2em}}
\newcommand{\funconvert}{\ttq{function\_convert}}
% --- Section styling (optional, but elegant) ---
\usepackage{titlesec}
\titleformat{\section}{\large\bfseries}{\thesection}{1em}{}
\titleformat{\subsection}{\normalsize\bfseries}{\thesubsection}{1em}{}


\usepackage{tcolorbox}
\tcbuselibrary{skins, breakable}
\tcbuselibrary{listings}

\newenvironment{maximabox}
  {\begin{tcolorbox}[
      breakable,
      enhanced,
      colback=black!2,
      colframe=black!50,
      boxrule=0.4pt,
      arc=2pt,
      left=6pt,
      right=6pt,
      top=6pt,
      bottom=6pt,
      before skip=10pt,
      after skip=10pt
    ]}
  {\end{tcolorbox}}

% reduce vertical space for display math
\setlength{\abovedisplayskip}{6pt}
\setlength{\belowdisplayskip}{6pt}
\setlength{\abovedisplayshortskip}{4pt}
\setlength{\belowdisplayshortskip}{4pt}

\DeclareMathOperator{\erf}{erf}
\DeclareMathOperator{\erfc}{erfc}
\DeclareMathOperator{\erfi}{erfi}
\DeclareMathOperator{\signum}{signum}
%magic for formatting for the bib
\usepackage{etoolbox}
\apptocmd{\thebibliography}{\raggedright}{}{}
% --- Microtypography ---
\usepackage[final,protrusion=true,expansion=true]{microtype}

% Optional: add the ToC to the PDF bookmarks if using hyperref
\usepackage{hyperref}
\hypersetup{
  colorlinks=true,
  linkcolor=black,
  urlcolor=blue,
  citecolor=black,
  pdfborder={0 0 0}
}

\usepackage[nameinlink]{cleveref}
% --- Metadata ---
\title{Writing Converters for the Function Convert Package}
\author{Barton Willis}
\date{\today}


\begin{document}
\maketitle
\tableofcontents


\section{Introduction and prerequisites}

Maxima’s \funconvert package~\cite{willis-function-convert} performs semantic function‑to‑function conversions on Maxima~\cite{maxima} expressions. This tutorial explains how to extend the package’s mathematical capabilities by writing your own converters. To follow this tutorial, you should be comfortable with:
\begin{itemize} 
\item using Maxima and the \funconvert package, and 
\item programming Maxima in Common Lisp. 
\end{itemize}
This guide is intended for users who want to create new converters. It concentrates on the converter‑writing process and includes only the API elements relevant to authoring converters.

To install this package, place the file \ttq{function\_convert.lisp} in a directory on Maxima’s search path and load it with  \ttq{load("function\_convert");}.  Eventually, this package might be migrated to 
Maxima's /src folder --- when this happens, loading the package will not be needed.

We begin with the basic usage of the \funconvert package and an overview of its main features.


\section{Usage and features}

The most basic example use of the package is a converter that replaces each subexpression of a Maxima expression of the form  $\erfc(X)$  by $1 - \erf(X)$, where  $\erfc$ and  $\erf$ are error functions (DLMF \S 7.2~\cite{DLMF}) and 
$X$ may be any expression. This converter is built into the \funconvert package. To use it, enter:
\begin{maximabox}
\begin{verbatim}
(%i1) function_convert(erfc = erf, erfc(x));
(%o1)                             1 - erf(x)
\end{verbatim}
\end{maximabox} 
\noindent  Of course, the input can be any expression, including nested calls to $\erfc$; for 
example
\begin{maximabox}
\begin{verbatim}
(%i2) function_convert(erfc = erf, x*erfc(x + erfc(z)));
(%o2)                     x (erf(erf(z) - x - 1) + 1)
\end{verbatim}
\end{maximabox} 
The first argument \ttq{erfc = erf} is called a \emph{conversion rule} (or simply a rule), 
the left-hand side of a rule is the \emph{source function} and the right-hand side \ttq{erf} is called  the
\emph{target function}. The rule \ttq{erfc = erf} does not literally
replace the source function with the target function, but it instead applies the 
identity $\erfc(x) = 1 - \erf(x)$ to an expression, replacing the left-hand side with 
the right-hand side. 

The target function is best thought of a mnemonic for recalling 
the identity $\erfc(x) = 1 - \erf(x)$, not as a literal function. Actually, the name of the 
target function could be any Maxima symbol; for example, the rule could be named 
\ttq{erfc = erfc\_to\_erf}.

This use of the infix operator \ttq{=} to define a rule is similar to its use in the Maxima function substitute, with one difference: here the operator \ttq{=}  indicates a semantic conversion rather than a literal renaming. One might argue that an arrow‑like operator such as \ttq{=>} would make this distinction clearer. Using \ttq{=}   instead of an arrow‑like symbol, however, keeps the notation consistent with substitute and preserves arrow‑like operators for other purposes.




\subsection{Algorithm and implementation}
\label{sec:algorithm}

The function \ttq{function-convert} recursively traverses the expression tree 
and replaces function calls according to well‑defined
rules.  In that sense it is straightforward code of a kind repeated
many times in Maxima.  It is not a pattern matcher or a general
rewrite engine.

Specifically, \ttq{function-convert} performs a \emph{preorder} (root–first, left‑to‑right) traversal of the expression tree: each node is visited before its children. When the traversal reaches an atom, it returns immediately and no converter is applied. For non‑atomic expressions, the converter is not applied at the moment the node is visited. Instead, the converter is applied only \emph{after} all arguments have been recursively converted.

This yields a hybrid order: the traversal is top--down, but converter
application is bottom--up.  For example, in the expression
\ttq{a + b*c}, the converter is applied first to \ttq{b*c} and only
then to the sum.

We can illustrate this evaluation scheme with a converter \ttq{algebraic = boxed} that
wraps every subexpression whose operator is \ttq{+}, \ttq{*}, or
\ttq{\^{}} in a consecutively numbered box.  This rule  is not built-in to the 
\funconvert package, but it can be written in just a few lines of code; for this code, 
see Section~\ref{sec:classkey}. Two examples:

\begin{maximabox}
\begin{verbatim}
(%i1) function_convert('algebraic = boxed, a+b*c);
(%o1) box(box(b*c,1)+a,2)

(%i2) function_convert('algebraic = boxed, a+b*(c+d));
(%o2) box(box(b*box(d+c,3),4)+a,5)
\end{verbatim}
\end{maximabox}
\noindent Two items to note: Maxima reorders commutative operands during canonicalization; in particular \ttq{c+d}
prints as \ttq{d+c}. Also since \ttq{algebraic} is a Maxima option variable (default \ttq{false}),
it is necessary to quote the source function for this converter. Notice that individual atoms are not boxed --- this is due to the policy that when the traversal reaches an atom, no converter is called and
the atom is returned immediately.



The package is implemented in Common Lisp.  It has been tested using
Clozure Common Lisp~\cite{ccl} version~1.13 and
SBCL~\cite{sbcl} version~2.4.7.


\subsection{Naming and the alias system}
\label{sec:alias_system}
We've seen that the name of the target function is best thought of an alias for the identity being 
applied.  The \funconvert package has an alias system allows the source function to be a
mnemonic as well.  This feature is useful for naming a converter for an identity 
such as 
\begin{equation}
\Gamma(z)\Gamma(1-z)=\uppi/\sin(\uppi z).
\label{gammaid}
\end{equation}
Here the left-hand side of this identity is a product so the source function for this identity must be multiplication (internally \ttq{mtimes}), not the gamma function. But the rule \ttq{”*” = sin} is
hard to remember and unnatural.  The alias system allows the converter for this identity to have a much more natural name. For the identity in Eq. (\ref{gammaid}), using the alias system we can 
name this rule \ttq{gamma = sin}, for example. Actually, this converter is built-in to the \funconvert package:

\begin{maximabox}
\begin{verbatim}
(%i1) function_convert(gamma = sin, gamma(x)*gamma(1-x));
(%o1) %pi/sin(%pi*x)

(%i2) function_convert(gamma = sin, gamma(2 + x)*gamma(1-x));
(%o2) gamma(1-x)*gamma(x+2)
\end{verbatim}
\end{maximabox}

A macro takes care of the messy details of defining an alias, along with other housekeeping duties. 
 
 
 \subsection{Built-in converters}

A converter may be marked as built-in. Built-in converters cannot be removed using any user-level function. For example, the converter \ttq{erfc = erf} is built-in; attempting to remove it with \ttq{delete\_converter} produces an error:
\begin{maximabox}
\begin{verbatim}
(%i1) delete_converter(erfc = erf);
Cannot delete built-in converter erfc = erf.
(%o1) done
\end{verbatim}
\end{maximabox}
 
 
 \subsection{Self-documentation feature}
 
 Every converter function should have a docstring.  The docstring is available to the user
 via the function \ttq{describe\_converter}; for example
\begin{maximabox}
  \begin{verbatim}
    (%i1) describe_converter(sinc = sin);
          Converter: sinc = sin
          Type: built-in
          Docstring: Convert sinc(x) into sin(x)/x.
    (%o1)                                done
  \end{verbatim}
\end{maximabox}
If a user-package properly includes a docstring for a converter, once the package is loaded, that documentation will be available via the function \ttq{describe\_converter} .  


\subsection{Class keys and class-based dispatch}

Converter functions often need to apply the same transformation to an entire
family of related operators. For example, a converter that rewrites
trigonometric functions into exponential functions should apply to
all six trigonometric functions. Writing six separate converters is tedious and 
error-prone. The \emph{class key} mechanism provides a solution.

\paragraph{Purpose of class keys} A \emph{class key} is a symbolic tag (such as \ttq{:trig} or
\ttq{:hyperbolic}) that represents a category of operators. The converter
system uses these tags to perform \emph{class-based dispatch}: a single
converter registered for a class key is automatically applied to every
operator belonging to that class.

Class keys classify operators, not expression types. For example, the class
\ttq{:algebraic} contains the operators \ttq{mplus}, \ttq{mtimes}, and \ttq{mexpt}.
If the head operator of an expression belongs to the class \ttq{:algebraic}, that does not imply that the expression itself is algebraic.

\paragraph{The class table} The mapping from class keys to operators resides in the 
global \ttq{*converter-class-table*}. This table is 

\begin{maximabox}
\begin{verbatim}
(defparameter *converter-class-table*
  '((:trig        . (%sin %cos %tan %sec %csc %cot))
    (:hyperbolic  . (%sinh %cosh %tanh %sech %csch %coth))
    (:inv_trig    . (%asin %acos %atan %asec %acsc %acot))
    (:inv_hyperbolic . (%asinh %acosh %atanh %asech %acsch %acoth))
    (:exp            . (mexpt))
    (:gamma_like     . (%gamma %beta %binomial %double_factorial
                        mfactorial $pochhammer))
    (:bessel  . (%bessel_j %bessel_y %bessel_i %bessel_k   %hankel_1 %hankel_2))
    (:algebraic . (mplus mtimes mexpt))
    (:inequation . (mequal mlessp mleqp mnotequal mgreaterp mgeqp
                    $notequal $equal))
    (:logarithmic . (%log)))
  "Mapping from class keys to lists of operator symbols.")
\end{verbatim}
\end{maximabox}
Each entry associates a class key with the list of operators that belong to
that class. The system assumes that \emph{an operator belongs to exactly one class}.
This ensures unambiguous dispatch: when the converter system encounters an
operator, it can determine its class uniquely and apply the appropriate
class-level converter.

\paragraph{How class dispatch works} When the converter system processes an expression, it identifies the head operator of the form. It then:
\begin{enumerate}[label=(\alph*), noitemsep, topsep=0pt]
  \item looks up the operator in \ttq{*converter-class-table*};
  \item determines the corresponding class key (e.g., \ttq{:trig});
  \item checks whether a converter is registered for that class;
  \item if so, applies that converter to the entire expression.
\end{enumerate}

\paragraph{Why class keys matter} Class keys provide:
\begin{itemize}
  \item \textbf{Abstraction}: one converter per conceptual family.
  \item \textbf{Extensibility}: adding a new operator to a class requires no
        changes to existing converters.
  \item \textbf{Uniformity}: guarantees consistent behavior across related
        operators.
  \item \textbf{Clarity}: the class table documents the structure of the
        operator space.
\end{itemize}

\subsection{Class--Key utilities}
\label{subsec:class-key-utilities}

There are two small utility functions that give convenient access to 
class--key information. 

\paragraph{From class key to operators.}
The function \ttq{class-table-members} returns the list of operator symbols
associated with a given class key.  For example,
\begin{maximabox}
\begin{verbatim}
> (class-table-members :bessel)
(%BESSEL_J %BESSEL_Y %BESSEL_I %BESSEL_K %HANKEL_1 %HANKEL_2)
\end{verbatim}
\end{maximabox}



\paragraph{From operator to class key.}
The inverse query is handled by \ttq{converter-class-of}, which scans the
class table and returns the unique class key containing the operator, or
\texttt{NIL} if the operator belongs to no class; for example

\begin{maximabox}
  \begin{verbatim}
    > (converter-class-of '%bessel_j)
     :BESSEL
  \end{verbatim}
\end{maximabox}

\subsection{Chaining multiple conversions}
To apply two or more conversions, put the converters into a Maxima list; for example

\begin{maximabox}
\begin{verbatim}
(%i1) function_convert(['sinc = 'sin, 'sin = 'exp], sinc(x)^2);
(%o1) -((%e^(%i*x)-%e^-(%i*x))^2/(4*x^2))
\end{verbatim}
\end{maximabox}
\noindent Converters are applied left-to-right. 

Chaining multiple conversions is handled by the function \ttq{function\_convert}; a writer of
a converter doesn't need to do anything to allow chaining to work. 

\subsection{Implicit dispatch via breadth-first search}

The converter system automatically discovers multistep conversion chains.
For example, there are rules $f = g$ and $g = h$, the system is able to implicitly apply the rule $f = h$.
In other words, there is no need to define a direct converter $f = h$.

This behavior is enabled by a breadth‑first search (BFS) over the converter graph ~\cite{clrs}. The search locates the shortest available path between two operators and composes the corresponding converters. When several shortest paths exist, the system selects one according to its internal traversal order; the choice among equal‑length paths is not user‑configurable. In such cases, the final result may depend on which shortest path is chosen.

Writers of converters only need to know that \funconvert uses BFS, not the internal details of its implementation.

\subsection{User-level converters}
\label{sec:register-converter}

The function \ttq{register\_converter} provides a user-level method to define a new converter using just the Maxima language.  A disadvantage to this method is that the class-key dispatch method is 
not available. 

A converter is specified by: (a) a name for the rule,  (b) the alias for the rule, and (c) a Maxima lambda expression implementing the conversion.  An optional documentation string may 
also be supplied.

Here is an example of a rule that expands noncommuative powers, but not noncommuative products
of sums.  We'll give the rule and its alias the same name:
\begin{maximabox}
\begin{verbatim}
(%i1)	register_converter("^^" = expand, "^^" = expand, 
      lambda([a,b], expand(a^^b)), "Expand noncommutative powers.");
(%o1)	done

(%i2)	function_convert("^^" =  expand, a.(b+c) + (1+x)^^3);
(%o2)	x^^3+3*x^^2+3*x+a.(c+b)+1
\end{verbatim}
\end{maximabox}
\noindent A converter defined this way is fully integrated into the built-in converter graph, and user-defined rules are discoverable by the BFS mechanism.  Also, \ttq{describe\_converter} works on 
user-defined rules:
\begin{maximabox}
\begin{verbatim}
(%i3)	describe_converter("^^" = expand);
Converter "^^""="expand
Type: "user-defined"
Docstring: "Expand noncommutative powers."
(%o3)	done
\end{verbatim}
\end{maximabox}

%%%%%%%%%%%%%%%%%%%%%%%
\subsection{Errors and debugging}

When a converter is redefined, Maxima will issue a warning, and when a converter
doesn't exist,  Maxima returns the input unchanged, but doesn't issue a warning; 
for example
\begin{maximabox}
\begin{verbatim}
(%i2) function_convert(f = g, a+b);
(%o2)                                b + a
\end{verbatim}
\end{maximabox}

When writing a new converter, if a converter doesn't trigger, the problem is likely due to either
a typographical error or some confusion between Maxima nouns and verbs. If a converter doesn't trigger 
as expected, try tracing the Common Lisp function \ttq{lookup-converter}. From there, you 
might be able to diagnose the problem.

\section{Writing converter functions}
A macro \ttq{define-function-converter} takes care of many of the housekeeping 
details in writing a converter, including naming the internal function that applies the rule, 
optionally marking the converter as built-in, and gathering the docstring 
for the converter to use a user-documentation. The macro also automatically handles 
the details needed to make the alias system work. 

The function \ttq{function\_convert} does the work of mapping
the converter over the expression tree; this allows the converter code to be relatively simple.
We'll begin by examining the source code of the \ttq{erfc = erf} converter.

\subsection{The $\erfc$ to $\erf$ converter}

 The source code for the $\erfc$ to $\erf$ converter is 
\begin{maximabox}
\begin{verbatim}
(define-function-converter (%erfc %erf) (op x)
    :builtin
  "Convert erfc(x) into 1 - erf(x)."
  (declare (ignore op))
  (let ((z (car x)))
    (sub 1 (ftake '%erf z))))
\end{verbatim}
\end{maximabox}
The arguments \ttq{\%erf \%erfc} are the source and target functions, 
respectively.  For this converter, the argument \ttq{op} is bound to the source
function \ttq{\%erfc}, but  this converter doesn't use this argument, so the
converter code should declare to ignore it.  The argument \ttq{x} is
a Common Lisp list of the argument to the function \ttq{\%erfc}.

The optional keyword argument  \ttq{:builtin} declares that this converter cannot be
deleted using user-level functions. To mark a rule as built-in, the keyword \ttq{:builtin}  must be placed before the docstring. For a non-built-in rule,  omit the  \ttq{:builtin} keyword.
The docstring is optional, but important because supplies the converter with documentation. 

The mathematical part of the code, that is
\begin{maximabox}
\begin{verbatim}
  (let ((z (car x)))
    (sub 1 (ftake '%erf z)))
\end{verbatim}
\end{maximabox}
\noindent implements the identity $\erfc(x) = 1 - \erf(x)$.  Notice that since \ttq{x} is a Common
Lisp list of the argument to \ttq{\%erfc}, it is necessary to apply \ttq{\%erf} to the car of \ttq{x}, and
not to \ttq{x} itself.  Alternatively, if you prefer, this bit of code can be changed to 
\begin{maximabox}
\begin{verbatim}
      (sub 1 (fapply '%erf x))
\end{verbatim}
\end{maximabox}

\subsection{Alias dispatch}

As we discussed before, both the source and the target functions of a converter can be given an alias.  This can be useful when the actual source function is a poor mnemonic for the converter name.

Again, the macro \ttq{define-function-converter} takes care of the housekeeping 
for defining an alias.  The best way to explain how define a converter that uses
alias dispatch is by an example.  Here we give a converter that expands explicit powers, but not 
products of sums.  The source function must be \ttq{mexpt}, but we'll make
\ttq{power = expand} an alias for \verb|”^” = expand|. This is how it is done:
\begin{maximabox}
\begin{verbatim}
(define-function-converter ((mexpt $expand) ($power $expand)) (op x)
  ($expand (fapply 'mexpt x)))
\end{verbatim}
\end{maximabox}
\noindent A rule can have at most one alias.

There are two ways to invoke this converter:
\begin{maximabox}
\begin{verbatim}
(%i1) function_convert("^" = expand, (a+b)*x + (1+x)^2);
(%o1) x^2+(b+a)*x+2*x+1

(%i2) function_convert(power = expand_powers, (a+b)*x + (1+x)^2);
(%o2) x^2+(b+a)*x+2*x+1
\end{verbatim}
\end{maximabox}



\subsection{Class key converter dispatch}
\label{sec:classkey}

Class-key converters allow a single definition to handle an entire family of
related operators, for example all trigonometric functions. The following example 
illustrates this idea using the trigonometric class \ttq{:trig}. Any operator listed under the
\ttq{:trig} class key in \ttq{*converter-class-table*} will be handled
by the same converter, without the need to define six separate rules. 

Specifically,  we'll show how to define a rule that rewrites all trigonometric functions in terms of sine
and cosine. For this converter, we finally have a reason for the \ttq{op} argument --- this
argument will be bound to one of the six trigonometric functions.  The
main code uses a \ttq{case} statement to handle each of the six cases. The code is

\begin{maximabox}
\begin{verbatim}
(define-function-converter (:trig $sin_cos) (op x)
  :builtin
  ;; Convert all six trigonometric functions to sin/cos form.
  (let ((z (first x)))
    (case op
      (%sin (ftake '%sin z))
      (%cos (ftake '%cos z))
      (%tan (div (ftake '%sin z) (ftake '%cos z)))
      (%sec (div 1 (ftake '%cos z)))
      (%csc (div 1 (ftake '%sin z)))
      (%cot (div (ftake '%cos z) (ftake '%sin z)))
      (t (ftake op z)))))
\end{verbatim}
\end{maximabox}
This converter is registered for the class key \ttq{:trig}, not for a particular
operator. When the dispatch system encounters an expression such as
\ttq{((\%tan) x)} or \ttq{((\%csc) x)}, it determines that the operator
belongs to the \ttq{:trig} class and invokes the converter above.

The body of the converter receives two arguments: the operator \ttq{op}
and its argument list \ttq{x}. The converter then rewrites each
trigonometric function into its sine-cosine form. 

We conclude this section with the converter \ttq{algebraic = boxed} introduced in Section~\ref{sec:algorithm}. This example combines class‑key and alias dispatch and serves primarily to illustrate the postorder application of converters; it is not intended to have mathematical significance.
\begin{maximabox}
\begin{verbatim}
(defmvar $box_counter 0)
(define-function-converter ((:algebraic $boxed) ($algebraic $boxed)) (op x)
   (ftake 'mlabox (fapply op x) (incf $box_counter 1))) 
\end{verbatim}
\end{maximabox}
\noindent This converter increments the global variable \ttq{box\_counter} each time it is applied. In general, converters should be pure functions without side effects.

\subsection{Converters for subscripted functions}

Maxima represents both the polygamma and the polylogarithm function as subscripted functions. For definitions of these functions, see DLMF  §5.15 and §25.12~\cite{DLMF}. Writing a converter for these functions, and for subscripted functions in general, requires understanding the internal structure of these expressions and using several higher‑level functions that extract their components. To illustrate this, we will build a converter that applies the polygamma half‑argument identity. This identity is
 \begin{equation}
 \psi_0\left(z\right)=\tfrac{1}{2} \psi_0\left(\tfrac{z}{2}\right)+
   \tfrac{1}{2} \psi_0\left(\tfrac{z+1}{2}\right)+\ln\left(2\right).
 \end{equation}
 To construct the converter, we first need to understand Maxima's internal structure for subscripted 
 functions.  For the polygamma function, internally  \(\psi_{n}(x)\)
 is \ttq{((MQAPPLY SIMP) ((\$PSI SIMP ARRAY) \$N) \$X)}. We say that \ttq{\$psi} is the \emph{name} of the expression; the Common Lisp list \ttq{(\$n)} is its \emph{subscript}, and the Common Lisp list \ttq{(\$x)} is its \emph{argument}.

The head operator of all subscripted functions is \ttq{mqapply}.  Thus,
the target function for a converter for a subscripted function must be \ttq{mqapply}. 
Thus, we will almost surely want to use an alias dispatch when dealing such cases.

Second, we need to know how to use several high-level functions that provide access the name, subscripts, and augments of  of a subscripted function. For any subscripted function  \ttq{e}, we have 
 \begin{enumerate}[label=(\alph*), itemsep=3pt, topsep=2pt, parsep=0pt]
  \item  \ttq{(subfunname e)} returns the name of the subscripted function.
    
  \item  \ttq{(subfunsubs e)} returns a Common Lisp list of the subscripts.
  
  \item \ttq{(subfunargs e)} returns a Common Lisp list of the arguments.
  
  \item \ttq{(subfunmake fn subs args)} constructs a subscripted function
  with name \ttq{fn}, subscripts \ttq{subs}, and arguments \ttq{args}.
 \end{enumerate}
 \noindent  With these functions we can define the converter as
 
 \begin{maximabox}
 \begin{verbatim}
 (define-function-converter ((mqapply $psi_half) ($psi $psi_half)) (op x)
 "Apply the polygamma half-argument identity; see http://dlmf.nist.gov/5.5.E8"
 (let* ((e (fapply op x)) 
        (fn (subfunname e))  (subscripts (subfunsubs e)) 
        (args (subfunargs e)) 
        (z (div (car args) 2)) 
        (n (car subscripts)))
  (cond ((and (eql n 0) (eq fn '$psi))
          (add (div (add (subfunmake  fn subscripts (list z))
                         (subfunmake fn subscripts (list (add (div 1 2) z))))
                 2) (ftake '%log 2)))
        (t e))))
   \end{verbatim}
 \end{maximabox}
In the \ttq{let*}, we first reconstruct the entire subscripted function, and then use
the functions \ttq{subfunname}, \ttq{subfunsubs}, and \ttq{subfunargs} to find the function name,
subscripts, and arguments.  After that, we check if the lone subscript is zero and if
the function name is \ttq{\$psi}, and if so, we
apply the half-argument identity; if not we return the entire subscripted function. 

Because the head operator of every subscripted function is \ttq{mqapply}, every converter
defined for a subscripted function must correctly handle every subscripted function, not just
the specific one that it is intended to handle. So when testing, be sure to check that your
converter handles a variety of subscripted functions correctly; for example
\begin{maximabox}
\begin{verbatim}
(%i1) function_convert(psi = psi_half, psi[0](x));
(%o1) (psi[0](x/2)+psi[0](x/2+1/2))/2+log(2)

(%i2) function_convert(psi = psi_half, li[2](x) + psi[1](x) + psi[0](x));
(%o2) li[2](x)+psi[1](x)+(psi[0](x/2)+psi[0](x/2+1/2))/2+log(2)
\end{verbatim}
\end{maximabox}

\subsection{An advanced converter}

In this section, we show to build a converter that rewrites explicit products of the form
$x  \signum(x)$ as $|x|$.   The converter does not rewrite 
\(\mathrm{signum}(x)\) itself, because the pattern \(x \signum(x)\) is 
not a subexpression of \(\signum(x)\); only explicit products matching that pattern
are transformed --- this is what makes writing the converter somewhat tricky.  The converter logic
needs to scan the arguments of each product and gather the terms that have $\signum$
as their head operator.  For each of these terms,  the converter does a rational substitution (\ttq{ratsimp}) of the form \ttq{ratsubst(abs(s), s*signum(s), e)} on the expression \ttq{e},
that must be a product. The definition is:

\begin{maximabox}
\begin{verbatim}
(define-function-converter ((mtimes mabs) (%signum mabs)) (op x)
  :builtin
  "Convert subexpressions of the form X*signum(X) into abs(X).  This converter
does not rewrite signum(X) itself, since X*signum(X) is not a subexpression
of signum(X); only explicit products matching that pattern are transformed."
  (let* ((e  (fapply op x))
         (ll (xgather-args-of e '%signum)))
    (dolist (lx ll)
      (let ((s (car lx)))
        (setq e ($ratsubst (ftake 'mabs s)
                           (mul s (ftake '%signum s))
                           e))))
    ($expand e 0 0)))
\end{verbatim}
\end{maximabox}

\subsection*{How the converter works}

\begin{enumerate}[label=(\alph*), noitemsep, topsep=0pt]
  \item The converter receives the head operator \ttq{op} and its
        argument list \ttq{x}.  The call \ttq{(fapply op x)}
        reconstructs the full expression  \ttq{e}.

  \item The function \ttq{xgather-args-of} scans \ttq{e}  for
        occurrences of \ttq{\%signum} and returns a list of matches.
        Each element of this list contains the argument \(X\) of a
        subexpression \(\signum(X)\). The \funconvert
        package defines the function \ttq{xgather-args-of}.

  \item For each such argument \(s\), the converter constructs the
        product \(s  \signum(s)\) and replaces it with
        \(|s|\) using \ttq{ratsubst}.  Only explicit
        products are rewritten; no algebraic inference is performed.

  \item After all substitutions, the expression is resimplified with a
        call to \ttq{expand}.
\end{enumerate}

This converter is intentionally conservative: it rewrites only the
literal pattern \(X  \signum(X)\), never the function
\(\signum\) itself.

%%%%%%%%%%%%%%
\section{Regression tests}

This section describes how to write regression tests for the \ttq{function\_convert} package 
and how to run them. 

\paragraph{Test file format} A regression test file is an ordinary Maxima batch file.  
Each test consists of \emph{exactly two lines}:
\begin{maximabox}
\begin{verbatim}
input;
expected output$
\end{verbatim}
\end{maximabox}
\noindent
The first line is the Maxima expression to evaluate, and the second line is the expected result.
Terminating the input line with a semicolon and the expected output with a dollar sign clarifies
the intent of each line, but the line terminators can be either a semicolon or a dollar sign, if you
prefer. A file can have as many tests as needed. Here is an example of one test:
\begin{maximabox}
\begin{verbatim}
function_convert('cos = 'exp, cos(x));
(%e^(%i*x)+%e^-(%i*x))/2$
\end{verbatim}
\end{maximabox}

\paragraph{Running the tests} If your tests are in a file \ttq{rtest\_my\_function\_convert}, to run the regression tests, enter the command:
\begin{maximabox}
\begin{verbatim}
batch(rtest_my_function_convert, 'test);
\end{verbatim}
\end{maximabox}
\noindent The command \ttq{batch} with an optional second argument of \ttq{test}, loads the test file, evaluates each input expression, compares the actual output with the expected output, and reports any mismatches.   A successful run prints no errors.

\paragraph{What to test} 

\begin{itemize}
\item bogus, unexpected (boolean values, strings, or \dots),  or invalid input,
  \item complex numbers,
  \item nested expressions and multiple occurrences the source function, and
  \item  converters applied to expressions containing unknown functions.
\end{itemize}

\paragraph{Writing good tests} A good regression test is minimal, documents the intended behavior of the converter system, and  isolates a single behavior. If a converter is based on an identity, each such converter should have an explicit test of the identity; here is an example that tests the identity 
$\erfc(x) = 1-\erf(x)$:
\begin{maximabox}
\begin{verbatim}
erf(x) + function_convert(erfc = erf, erfc(x));
1$
\end{verbatim}
\end{maximabox}

\section{Best practices for writing converters}

 The guidelines below help ensure that new converters integrate smoothly into
the \funconvert system.

\begin{itemize}

  \item \textbf{Base converters on identities.}
        A converter should be based on a mathematical identity.
        It should not change the meaning of an expression. If a transformation is only valid under certain
        assumptions, the converter should check them explicitly or decline
        to apply the transformation. It is discouraged, but possible, to
        write a converter that, for example, replaces every product by a sum.

  \item \textbf{Include a docstring.}
        Each converter should have a docstring so that \funconvert is
        self-documenting. If a converter requires a domain restriction or
        mathematical convention, document it. Clear documentation prevents
        misuse and helps future contributors maintain the system.

  \item \textbf{Use clear, descriptive names.}
        Converter names or aliases should reflect the mathematical
        transformation they perform. This helps users understand the
        converter registry and simplifies debugging.

  \item \textbf{Keep converters small and focused.}
        A converter should perform one clear transformation, such as
        rewriting \(\erfc(x)\) as \(1 - \erf(x)\) or converting
        \(\tan(x)\) to \(\sin(x)/\cos(x)\). Small, focused converters
        compose more reliably and are easier to test and debug.

  \item \textbf{Dispatch on functions, not patterns.}
        Converters operate on function calls, not syntactic patterns. Let
        the converter receive the arguments directly and compute the
        replacement semantically. This avoids brittle rules and reduces
        unintended matches.

  \item \textbf{Ensure termination.}
        Converters should not create expressions that trigger themselves
        again unless the transformation is guaranteed to converge.
        
\item \textbf{No side effects.} Converters should be functions with no side effects.

  \item \textbf{Write regression tests.}
        Each converter should have regression tests covering typical
        inputs, edge cases, and expressions where the converter should
        \emph{not} apply. Tests help maintain stability as the system
        evolves.

\end{itemize}


The example that follows illustrates why a converter should be based on an 
identity.  If it is not, the results produced by the traversal are
predictable but mathematically meaningless.  Say we attempt
to write a converter that returns an antiderivative; our code is
\begin{maximabox}
\begin{verbatim}
;; This demonstrates that converters should be based on identities.
;; When they are not, the results are predictable but useless.
(define-function-converter ((:algebraic $integrate)
                            ($algebraic $integrate)) (op x)
  (let ((e (fapply op x)))
    ($integrate e '$x)))
\end{verbatim}
\end{maximabox}

Applied to an expanded polynomial, the converter appears to behave
correctly:

\begin{maximabox}
\begin{verbatim}
(%i1) function_convert('algebraic = integrate, x^2+1);
(%o1) x^3/3 + x
\end{verbatim}
\end{maximabox}

This happens because the converter maps over the arguments of the sum
and encounters only atoms.  By the traversal rules, it returns each
atom immediately, so the converter ultimately receives the entire
expression \ttq{x\^{}2+1}.  The call to \texttt{integrate} therefore
produces a proper antiderivative.

In contrast:

\begin{maximabox}
\begin{verbatim}
(%i2) function_convert('algebraic = integrate, x);
(%o2) x
\end{verbatim}
\end{maximabox}
\noindent Here the return-immediately-for-an-atom rule applies at the root, so
the converter is never invoked.  The result is simply the original
expression, not an antiderivative.

A more striking illustration comes from comparing an unexpanded product
with its expanded form:
\begin{maximabox}
\begin{verbatim}
(%i3) function_convert('algebraic = integrate, (x+1)*(x+2));
(%o3) (6*x^5+45*x^4+80*x^3)/120

(%i4) function_convert('algebraic = integrate, expand((x+1)*(x+2)));
(%o4) (5*x^3)/6 + 2*x
\end{verbatim}
\end{maximabox}
\noindent In \texttt{(\%o3)}, the converter returns an antiderivative of a 
product of antiderivatives of the factors of the input, not an antiderivative
of the input. Of course, both results are follow directly from the traversal rules, 
but only the latter corresponds to a meaningful identity.

This example underscores the principle that converters should express
identities.  When they do not, the traversal produces consistent results, but 
those results need not have any mathematical significance.

\section{Miscellaneous}

We end with a few comments about the \funconvert package that are not directly related to the process of writing converter functions.

\subsection {Motivation} 

Many systems (Maple~\cite{maple-convert}, Mathematica~\cite{mathematica-rules}, and SymPy~\cite{sympy-rewrite}) provide built‑in expansions or rewrite mechanisms, but Maxima uses an alphabet soup of 
functions that perform semantic function‑to‑function conversions; examples include  \ttq{makefact}
and \ttq{makegamma}. In other cases, transformations are controlled by option variables, for example, 
\ttq{expintrep}. 

These names are easy to forget and are not always easy to locate in the user documentation. 
The \funconvert package may offer a simple, uniform, and user‑extensible way to perform such conversions.

\subsection{Using  \ttq{defrule} as an alternative to \funconvert}

Maxima's pattern-based \ttq{defrule} tool is an alternative to using \funconvert. A simple
example: 
\begin{maximabox}
\begin{verbatim}
(%i1) matchdeclare(aa,true)$

(%i2) defrule(sinc_rule, sinc(aa), sin(aa)/aa);
(%o2) sinc_rule:marrow(sinc(aa),sin(aa)/aa)

(%i3) apply1(sinc(sinc(x)), sinc_rule);
(%o3) (x*sin(sin(x)/x))/sin(x)
\end{verbatim}
\end{maximabox}
This method works well, especially for single‑use function‑to‑function conversions. But there are several downsides to the \ttq{defrule} package.

First, a \ttq{kill(all)} removes all rules defined by \ttq{defrule}, and that makes the rule system fragile and difficult to use for anything beyond disposable transformations. Changing Maxima to allow rules that cannot be removed by \ttq{kill(all)} is, of course, a small matter of programming \cite{nardi-small-matter}, but it would violate a long‑standing design principle: users must always be able to return the system to a clean, rule‑free state.  In contrast, rules that are built-in to \funconvert are not removed by \ttq{kill(all)}.

Second, the \ttq{defrule} package still depends on a collection of auxiliary functions, and unlike
\funconvert, there the \ttq{defrule} package does not have a self-documentation feature.

Third, and most importantly, it is difficult to use \ttq{defrule} to define a rule that must match two or more terms in a sum or product. The \funconvert package has several converters that do exactly this. These converters often rely internally on \ttq{ratsubst} to perform semantic substitutions --- something that is difficult for a rule defined by \ttq{defrule} to accomplish. An example is the built‑in rule that applies the identity \(\Gamma(z)\Gamma(1-z)=\uppi/\sin(\uppi z)\) . The rule must scan a product and look for a pair of gamma functions whose arguments sum to one. Doing this with \ttq{defrule} is difficult to impossible, but it can be done using \funconvert as we saw in Section~\ref{sec:alias_system}.

We end this comparison with one possible advantage of the \ttq{defrule} package over \funconvert. The \ttq{defrule} package provides several mechanisms, including \ttq{apply1}, \ttq{apply2}, and \ttq{applyb1}, that give the user control over the order in which rules are applied, whereas the \funconvert package uses a fixed scheme that is not under user control.


\section{Glossary}

\begin{description}

\item[atom]  
Any expression for which the Maxima predicate \ttq{mapatom} returns \ttq{true}.  Maxima's
atoms are numbers, symbols, strings, and subscripted variables (for example, \ttq{x[1]}).

\item[bottom--up rewriting]  
The discipline in which converters are applied only after all children have been recursively converted.  
Although the traversal is preorder, rewriting is effectively postorder.

\item[canonicalization]  
Maxima’s internal reordering and normalization of expressions.  
For commutative operators, operands may be reordered (e.g., \texttt{c+d} prints as \texttt{d+c}).  
Converters always receive expressions in canonical form.

\item[converter]  
A pure function associated with a specific operator or a set of operators.
It receives an expression whose children have already been converted, and returns either a rewritten expression or the original expression unchanged.

\item[expression tree]  
The canonical Maxima representation of an expression, viewed as a rooted tree whose internal nodes are operators and whose leaves are atoms.

\item[operator]  
The function symbol at the head of a non-atomic expression.  
For example, in \ttq{x+y+z}, the operator is \ttq{+}.

\item[postorder]  
A traversal scheme in which each node is visited after all of its children have been visited.

\item[preorder traversal]  
A traversal in which each node is visited before its children (root--first, left-to-right).  
\ttq{function-convert} uses preorder visitation but applies converters only after all children have been recursively converted.

\item[pure function]  
A function with no side effects and no dependence on external state.  
Converters must be pure: they must not mutate their input or rely on evaluation.

\item[registry]  
The internal hash table mapping source operators to their associated converters.  
Lookup is performed during traversal; if no converter is registered for an operator, the expression is returned unchanged.

\item[source function]  
The operator for which a converter is registered.  
If the source function is a noun or option variable, it must be quoted when passed to \texttt{function-convert}.

\item[target function]  
The operator produced by a converter.  
A converter may return an expression with the same operator or a different one.

\end{description}

\paragraph*{Acknowledgements} The author thanks Robert Dodier, Richard Fateman, Stavros Macrakis, and Raymond Toy for their valuable guidance and constructive discussions on
the Maxima mailing list~\cite{maxima-list}. Their insights and
encouragement contributed significantly to the development and
refinement of this work.



\bibliographystyle{IEEEtran}
\bibliography{references}

\clearpage




\paragraph*{About the author}
This manual was written by Barton Willis, Professor Emeritus of Mathematics at the University of Nebraska at Kearney, a long‑time contributor to the Maxima computer algebra system, and the author of the \funconvert package. He has worked in symbolic computation, mathematical software, and technical documentation for many years.


\paragraph*{Colophon} This manual was typeset using the \texttt{LuaLaTeX}  engine with the
\texttt{extarticle} document class. The body text is set in a serif typeface chosen for
clarity in technical prose. Code fragments and Maxima sessions
are rendered in a monospaced font with consistent verbatim handling. Bibliographic data were managed using \texttt{bibtex} using the \texttt{IEEEtran} style.

All examples were executed using Maxima from the current development branch together with the \funconvert package also from the current development branch.

This manual is intended as a long-term reference for users and
developers who wish to extend the mathematical capabilities of the
\funconvert package by writing custom converters.

\begin{flushleft}
\textcopyright 2026 Barton Willis\\
Version 1.0\\
Portions adapted from the GitHub project README~\cite{willis-function-convert} with permission.\\[1ex]
Permission is granted to copy and distribute this document, in print or electronically, for any purpose, provided that the content is not
 modified, and this notice is preserved.
\end{flushleft}

\end{document} 

Sections I might include 

- Quick Start
- Conceptual Overview

- Error Handling and Debugging
- Regression tests
- Appendix: Built‑In Converters
- API of all public functions

-Error Handling and Debugging:
- common mistakes,
- how to interpret converter warnings,
- debugging tools or strategies,
- how to inspect the converter registry.



